---
title: "Basic Data Visualization with ggplot"
author: "Md Johirul Islam"
date: "2/26/2022"
output:
  html_document: default
  pdf_document: default
---

```{r library}
library(tidyverse)
library(socviz)
```

We will use "gapminder" library for the datasets. At first, we need to
install it, and then load it to the console. 

```{r gapminder}

library(gapminder)

gapminder #load the dataset

```

*Make a scatterplot*
```{r scatterplot}
p <- ggplot(data = gapminder, 
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()

```

*More details of ggplot*

#In ggplot, the overall type of plot is called a geom. Each geom has a 
function that creates it. For example, geom_point() makes scatterplots,
geom_bar() makes barplots, geom_boxplot() makes boxplots, and so on. 
You combine these two pieces, the ggplot() object and the geom, 
by literally adding them together in an expression, using the '+' symbol.

# In ggplot, mappings are specified using the aes() function.

*we include coordinates and scales by "coord_cartesian()" and " scale_x_log10()"*

#Lables and guides

```{r labs and guides}
p + labs(x ="Log GDP",
         y = "Life Expectancy",
         title = "A Gapminder plot")
```



*Build your plots layer by layer*

#Steps:
1. Tell the ggplot() function what our data is.         
2. Tell ggplot() what relationships we want to see.     
For convenience we will put the results of the 
first two steps in an object called p.
3. Tell ggplot how we want to see the relationships in our data.Choose a geom. 
4. Layer on geoms as needed, by adding them to the p object one at a time.
5. Use some additional functions to adjust scales, labels, tick marks, titles. We will learn more about some of these functions shortly.             

```{r geom smoother}
p + geom_smooth()   #use smoother graph

#use both scatterplot and smoother
p + geom_point() + geom_smooth()
```

```{r geom smooth for OLS}
p + geom_point() + geom_smooth(method = "lm")
```

#to make GDP normally distributed, we can scale the x-axis by taking log

```{r scaling x }
p + geom_point() + geom_smooth(method = "gam") + 
  scale_x_log10()

#we want top labell x-axis with dollar value, then
p + geom_point() + geom_smooth(method = "lm") + 
  scale_x_log10(labels = scales::dollar) #we use "::" to extract a function 
                                          #from a library not loaded.
```
 

#Mapping aesthetics vs. Settings them.#

Suppose, we want to set color. We do based on a categorical variable. 

```{r color setting by variable continent}

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap,
                          y = lifeExp,
                          color = continent))  #will color based on "continent"
p + geom_point() +
    geom_smooth(method = "loess") +
    scale_x_log10(labels = scales::dollar)
```
*Mapping the continent variable to the color aesthetic,*
*and correcting the error bars using the fill aesthetic.*

```{r correction for fill aesthetic}
p<- ggplot(data = gapminder, 
            mapping = aes(x  = gdpPercap,
                          y = lifeExp,
                          color = continent,
                          fill = continent))
p + geom_point() + 
    geom_smooth(method = "loess") +
    scale_x_log10(labels = scales::dollar)

```




If we want to color all points, we do the following: 

```{r color all points}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap,
                          y = lifeExp,
                          color = "purple")) #aes fn thinks purple as a variable
                                            #hence, it will create a new colum.
p + geom_point() +
    geom_smooth(method = "loess") +
    scale_x_log10()

```
The aes() function is for mappings only. Do not use it to change properties to a particular value. If we want to set a property, we do it in the geom_ we are using, and outside the mapping = aes(...) step. Try this:

*so for color, we use "geom_point", not aes in mapping*

```{r color with geom}
 p <- ggplot(data = gapminder, 
             mapping = aes(x = gdpPercap,
                           y = lifeExp))
p + geom_point(color = "purple") + 
    geom_smooth(method = "loess") + 
    scale_x_log10(labels = scales::dollar)
```

To set the color and size of the line, we use "geom_smooth" fn. 
```{r color and size of ols}
p + geom_point(alpha = 0.3) + #alpha controls the opacity of all points. 
  geom_smooth(color = "yellow",
              se  = F, #meaning that standard error is not known
              size = 3,
              method = "lm") + 
  scale_x_log10(labels = scales::dollar)
```

Let us make the complete graph with title, subtitle, caption. 
```{r complete graph}
 p <- ggplot(data = gapminder, 
             mapping = aes(x = gdpPercap,
                           y = lifeExp))

p + geom_point(alpha = 0.3, color = "purple") + 
    geom_smooth(color  = "red", size = 3, se = F, method = "lm") + 
    scale_x_log10(labels = scales::dollar) + 
    labs( x = "GDP per capita", y = "Life Expectancy",
          title = "Economic growth vs. Life Expectancy", 
          subtitle = "Data points are country-years", 
          caption = "Source: Gapminder")
```



**Aesthetics can be mapped per geom**

If we want one line, but need all points colored based on "continent". We use
"geom_point" fn. 

```{r one line, but colored points by continent}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap,
                          y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + 
    geom_smooth(method = "loess") + 
    scale_x_log10(labels = scales::dollar)
```
*Mapping a continuous variable to color.*
We can mapped a continuous variable such as "population" as follows: 

```{r mapping continous variable, fig.width = 9, fig.height = 9} 
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap,
                          y = lifeExp))
p + geom_point(mapping = aes(color = log(pop))) + 
    scale_x_log10(labels = scales::dollar) 
#save the single figure
ggsave(filename = "Mapping continuousV.png") #will save the file in png

```

*Save your work* 

You can also pass plot objects to ggsave(). For example, we can put our recent plot into an object called p_out and then tell ggave() that we want to save that object.

```{r using ggsave for direct object}
p_out <- p + geom_point() +
    geom_smooth(method = "loess") +
    scale_x_log10()

ggsave("my_figure.pdf", plot = p_out)

```


