---
title: "More on ggplot"
author: "Md Johirul Islam"
date: "2/28/2022"
output:
  pdf_document: default
  html_document: default
---

**Grouped data and the group aesthetic**

First, we need to load libraries
```{r load required libraries}
library(tidyverse)
library(socviz)
library(gapminder)

#load the data
gapminder
```

We can draw geom_line to show the trajectory of life expectancy. 

```{r drawing geom_line}
p<- ggplot(data = gapminder,
           mapping = aes(x = year,
                         y = lifeExp))
p+ geom_line()
```
Something bizarre happens. Here, we can use the group aesthetic to tell ggplot explicitly about this country-level structure.
```{r group year by country}
p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = lifeExp))
p + geom_line(aes(group = country))
```
*The group aesthetic is usually only needed when the grouping information you need to tell ggplot about is not built-in to the variables being mapped. For example, when we were plotting the points by continent, mapping color to continent was enough to get the right answer, because continent is already a categorical variable, so the grouping is clear. When mapping the x to year, however, there is no information in the year variable itself to let ggplot know that it is grouped by country for the purposes of drawing lines with it. So we need to say that explicitly.*


**Facet to make small multiples**

 We will use facet_wrap() to split our plot by a categorical variable since the
 plot looks messy. In this case, we split our plot by continent. 
 
```{r split plot by facet_wrap fn}
p<- ggplot(data = gapminder,
           mapping = aes(x = year,
                         y = lifeExp))
p + geom_line(aes(group = country)) + facet_wrap(~ continent)

```
*We can also use the ncol argument to facet_wrap() to control the number of columns used to lay out the facets. Because we have only five continents it might be worth seeing if we can fit them on a single row (which means we will have five columns)*

So, full figure is as follows: 

```{r complete figure, fig.height=8, fig.width=25}
p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = lifeExp,
                          color = continent,
                          fill = continent))
p+ geom_line(aes(group = country), color = "grey70") + 
  geom_smooth(method = "loess", size = 1.1, se = F) + 
  scale_y_log10() + 
  facet_wrap(~ continent, ncol = 5) + 
  labs(x = "Year",
       y = "Life Expectancy",
       title = "Life Expectancy on five continents")

```
*Facets can be more complex than this. For instance, you might want to cross-classify some data by two categorical variables. In that case you should try facet_grid() instead.*



**Faceting on two categorical variables.**
Here, we will use gss_sm dataset.

```{r load data}
gss_sm

#get a glimpse of the data
glimpse(gss_sm)

```
We use R's formula notation in the facet_grid function to facet sex and race. This time, because we are cross-classifying our results, the formula is two-sided: facet_grid(sex ~ race).

```{r facet for two categorical variable}
p <- ggplot(data = gss_sm,
            mapping = aes(x = age, y = childs))
p + geom_point(alpha = 0.2) +
    geom_smooth() +
    facet_grid(sex ~ race)

```
*We are not limited to two-way comparison. Further categorical variables can be added to the formula, too, (e.g. sex ~ race + degree) for more complex multi-way plots*



#Geoms can transform data#

To transfrom data we use stat_smooth fn.

**A bar chart.**
We plot bar chart by geom_bar fn. 
```{r bar chart}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))

p + geom_bar()

```
This bar chart gives the count on the vertical axis. Suppose, we want frequency 
rather than count.

*If we want a chart of relative frequencies rather than counts, we will need to get the prop statistic instead. When ggplot calculates the count or the proportion, it returns temporary variables that we can use as mappings in our plots. The relevant statistic is called ..prop.. rather than prop. To make sure these temporary variables won't be confused with others we are working with, their names begin and end with two periods. (This is because we might already have a variable called count or prop in our dataset.) So our calls to it from the aes() function will generically look like this: <mapping> = <..statistic..>. In this case, we want y to use the calculated proportion, so we say aes(y = ..prop..).*

```{r incorrect proportion}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))

p + geom_bar(mapping = aes(y = ..prop..))

```
The resulting plot is still not right. We no longer have a count on the y-axis, but the proportions of the bars all have a value of 1, so all the bars are the same height.
This is a grouping issue again. In this case, we need to tell ggplot to ignore the x-categories when calculating denominator of the proportion, and use the total number observations instead. To do so we specify group = 1 inside the aes() call. The value of 1 is just a kind of "dummy group" that tells ggplot to use the whole dataset when establishing the denominator for its prop calculations.

```{r bar chart with correct proportion}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))
p + geom_bar(mapping = aes( y = ..prop.., group = 1))

```

*making table from a categorical variable*

```{r making table for categorical variable}
table(gss_sm$religion)
```
We can make a bar plot for religion (categorical variable) with color and fill

```{r bar plot with color}
p<- ggplot(data = gss_sm,
           mapping = aes(religion, color = religion)) 
p + geom_bar()

```

```{r bar plot with fill}
p<- ggplot(data = gss_sm,
           mapping = aes(religion)) 
p + geom_bar(mapping = aes(x = religion, fill = religion))

#since legends are redundant, we can remove them. 
p + geom_bar(mapping = aes(x = religion, fill = religion)) +
    guides(fill = F) #use guides fn to remove legend, F means remove legends

```


#Frequency plots the slightly awkward way#

*cross-classify two categorical variables. W do it by ataked barplot*

This is the graphical equivalent of a frequency table of counts or proportions. 
Using the GSS data, for instance, we might want to examine the distribution of 
religious preferences within different regions of the United States.

```{r staked bar plot for cross-category}
p<- ggplot(data = gss_sm,
           mapping = aes(x = bigregion,
                         fill = religion))
p + geom_bar()

```

Now, we plot staked barplot with relative frequency
```{r staked barplot with frequency}
p<- ggplot(data = gss_sm,
           mapping = aes(x = bigregion,
                         fill  = religion))
p + geom_bar(position = "fill")

```
**A first go at a dodged bar chart with proportional bars**

```{r dodged bar 1st try}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion,
                          fill = religion))
p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop.. )) # to make y-axis as proportion

```

there seems to be an issue with the grouping. When we just wanted the overall
proportions for one variable, we mapped group = 1 to tell ggplot to calculate 
the proportions with respect to the overall N. In this case our grouping 
variable is religion, so we might try mapping that to the group aesthetic.

```{r dodged bar with grouping}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion, fill = religion))
p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop.., group = religion))

```
**Faceting proportions within region.**
We can plot different region in different panels bby faceting the plot.

```{r faceting barplot}
p <- ggplot(data = gss_sm,
            mapping = aes(x = religion))

p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop.., group = bigregion))  + 
    facet_wrap(~ bigregion, ncol = 2) #make it as four diff plot

```


#Histograms and density plots.#

A histogram is a way of summarizing a continuous variable by chopping it up 
into segments or "bins" and counting how many observations are found within each 
bin. Because we are summarizing a continuous variable using a series of bars,
we need to divide the observations into groups, or bins, and count how many are
in each one. By default, the geom_histogram() function will choose a bin size 
for us based on a rule of thumb.

```{r histograms with default bin size}
p <- ggplot(data = midwest,
              mapping = aes(x = area))

p + geom_histogram()

```

```{r histogram with custom bin size}
p <- ggplot(data = midwest,
              mapping = aes(x = area))
p + geom_histogram(bins = 10)

```

While histograms summarize single variables, it's also possible to use several
at once to compare distributions. We can facet histograms by some variable of 
interest, or as here we can compare them in the same plot using the fill mapping.

```{r two histogram in one plot}
oh_wi <- c("OH", "WI")

p <- ggplot(data = subset(midwest, subset = state %in% oh_wi), #to subset data
            mapping = aes(x = percollege, fill = state))
p + geom_histogram(alpha = 0.4, bins = 20)

```
*We subset the data here to pick out just two states. To do this we create a* 
*character vector with just two elements, "OH" and "WI". Then we use the*
*subset() function to take our data and filter it so that we only select*
*rows whose state name is in this vector. The %in% operator is a convenient* 
*way to filter on more than one term in a variable when using subset().*


**Kernel Density Plot**

We draw Density plot by 'geom_density' fn. 
```{r Kernel Density Plot}
p <- ggplot(data = midwest,
            mapping = aes(x = area))
p + geom_density()

```

We can use color and fill for this plot to compare more than one distributions. 

```{r more than one density plot }
p <- ggplot(data = midwest,
            mapping = aes(x = area,
                          fill = state,
                          color = state))
p + geom_density(alpha = 0.3)

```

If you want to make the baselines of the density curves go away, you can use geom_line(stat = "density") instead.

```{r without baseline density curves}
p <- ggplot(data = midwest,
              mapping = aes(x = area,
                            fill = state,
                            color = state))
p + geom_line(stat = "density")

```


Just like geom_bar(), the count-based defaults computed by the stat_ functions 
used by geom_histogram() and geom_density() will return proportional measures 
if we ask them. For geom_density(), the stat_density() function can return its 
default ..density.. statistic, or ..scaled.., which will give a proportional 
density estimate. It can also return a statistic called ..count.., which is 
the density times the number of points. This can be used in stacked density plots.

```{r scaled density}
p <- ggplot(data = subset(midwest, subset = state %in% oh_wi),
            mapping = aes(x = area, fill = state, color = state))
p + geom_density(alpha = 0.3, mapping = (aes(y = ..scaled..)))  #scaled density

```


###Avoid transformations when necessary###

When possible, we avoid transformation bcz sometimes our data is already in 
summary table. Let's consider the codes below. we do not need the services of any stat_ functions that geom_bar() would normally call. We can tell geom_bar() not to do any work on the variable before plotting it. To do this we say stat = 'identity' in the geom_bar() call. We'll also move the legend to the top of the chart.

```{r no transformation}
#load data
titanic

p <- ggplot(data = titanic,
            mapping = aes(x = fate, y = percent, fill = sex)) #make fill by sex

p+ geom_bar(position = "dodge", stat = "identity")  + #don't transform data to table 
            theme(legend.position = "top")   #make legend on the top of fig

```

For convenience ggplot also provides a related geom, geom_col(), which has exactly the same effect but assumes that stat = "identity". We will use this form in future when we don't need any calculations done on the plot. 


###Bar chart to compare groups for a threshold.###


This allows us to do things like, for example, plot a flow of positive and negative values in a bar chart. This sort of graph is an alternative to a line plot and is often seen in public policy settings where changes relative to some threshold level or baseline are of interest. For example, the oecd_sum table in socviz contains information on average life expectancy at birth within the United States, and across other OECD countries. We will plot the difference over time, and use the hi_lo variable to color the columns in the chart. 

```{r Using geom_col() to plot negative and positive values in a bar chart}

#load data
oecd_sum

p <- ggplot(data = oecd_sum,
            mapping = aes(x = year, y = diff, fill = hi_lo ))
p+ geom_col() + guides(fill = F) + #tells ggplot to drop the unnecessary legend
      labs(x = NULL, y = "Difference in years",
          title = "Life expectancy of US", 
          subtitle = "Difference between US and OECD in life expectancy",
          caption = "Data: OECD")

``` 












